---
title: "Workshop: Neural Networks with R - TEACHERS EDITION"
output: html_document
---

_**This is the completed R Markdown file to use only after the end of the workshop.**_

## Setup

The `saturn-rstudio-tensorflow` image has the required libraries
preinstalled--you just need to import them.

```{r setup_libraries}
library(dplyr)
library(readr)
library(stringr)
library(purrr)
library(tidyr)
library(keras)
```

Define what characters can be used for the pet names, and how far back the
neural network should look when generating them.

```{r setup_lookup_tables}
character_lookup <- data.frame(character = c(letters,".","-"," ","+"))
character_lookup[["character_id"]] <- 1:nrow(character_lookup)

max_length <- 10
num_characters <- nrow(character_lookup) + 1
```

Finally, download the raw data and format it into a table. This formats the
names and removes names with unusual characters. You don't really need to
understand what's happening 

```{r setup_cleaned_data}
data_url <-
  "https://saturn-public-data.s3.us-east-2.amazonaws.com/pet-names/seattle_pet_licenses.csv"
pet_data <- 
  read_csv(data_url, 
           col_types = cols_only(`Animal's Name` = col_character(),
             Species = col_character(),
             `Primary Breed` = col_character(),
             `Secondary Breed` = col_character())) %>%
  rename(name = `Animal's Name`,
         species = `Species`,
         primary_breed = `Primary Breed`,
         secondary_breed = `Secondary Breed`) %>%
  mutate_all(tolower) %>%
  filter(!is.na(name),
         !is.na(species),
         name != "",
         !str_detect(name,"[^ \\.-[a-zA-Z]]")) %>%
  sample_n(n())
```

## Model 1 - generate pet names

### Create training data

Next, we take the downloaded data and modify it so it's ready for the model. 
First we add stop characters to signify the end of the name ("+"), expand
the names into sub-sequences so we can predict each character in the name.

```{r model_1_subsequence_data}
subsequence_data <-
  pet_data %>%
  mutate(accumulated_name =
           name %>%
           str_c("+") %>%
           str_split("") %>%
           map( ~ purrr::accumulate(.x,c))
         ) %>%
  select(accumulated_name) %>%
  unnest(accumulated_name) %>%
  sample_n(n()) %>%
  pull(accumulated_name)
```

Then we make all the sequences the same length by truncating or padding them
so they can be a matrix. We also 1-hot encode the data.

```{r model_1_data_matrix}
text_matrix <-
  subsequence_data %>%
  map(~ character_lookup$character_id[match(.x,character_lookup$character)]) %>% 
  pad_sequences(maxlen = max_length+1) %>% 
  to_categorical(num_classes = num_characters) 
```

Once that's complete, we split the data into the 3D-matrix of model input (X) and
matrix of targets (y). We'll make the X matrix of all the letters in each row except
the last. The y matrix will be the last character (since we want to predict it).

```{r model_1_split_x_y}
x_name <- text_matrix[,1:max_length,]
y_name <- text_matrix[,max_length+1,]
```

### Create the model

Next we define the layers of the Keras model. This model has 2 LSTM layers to
find the patterns in the names, a dense layer to predict a value for each
possible next character, and a softmax activation to turn
those values into probabilities. Since this is a multiclass classification problem,
the loss is categorical cross-entropy.

```{r model_1_define_model}

# we'll fill this in

```

### Train the model

Once the model is defined, we can train it. Notice that if you run this
function again it won't train a new model, it'll fit more to the existing one (!)

```{r model_1_train_model}

# we'll fill this in

```

### Generate names

The function below generates a pet name using the trained model. It generates
characters and adds them to a running list of a name. It chooses each character
by pulling from the probability distribution from the model output.

```{r model_1_generate_names}
generate_name <- function(model, character_lookup, max_length, temperature = 1){
  choose_next_char <- function(preds, character_lookup, temperature){
    preds <- log(preds)/temperature
    exp_preds <- exp(preds)
    preds <- exp_preds/sum(exp(preds))
    next_index <- which.max(as.integer(rmultinom(1, 1, preds)))
    character_lookup$character[next_index-1]
  }
  
  in_progress_name <- character(0)
  next_letter <- ""
  
  while(next_letter != "+" && length(in_progress_name) < 30){
    previous_letters_data <- 
      list(in_progress_name) %>%
      map(~character_lookup$character_id[match(.x,character_lookup$character)]) %>%
      pad_sequences(maxlen = max_length) %>%
      to_categorical(num_classes = num_characters)
    
    next_letter_probabilities <- predict(model,previous_letters_data)
    
    next_letter <- choose_next_char(next_letter_probabilities,
                                    character_lookup,
                                    temperature)
    
    if(next_letter != "+")
      in_progress_name <- c(in_progress_name,next_letter)
  }
  
  raw_name <- paste0(in_progress_name, collapse="")
  
  capitalized_name <-gsub("\\b(\\w)","\\U\\1",raw_name,perl=TRUE)
  
  capitalized_name
}
```

You can then generate a name by calling the function:

```{r model_1_generate_one_name}
generate_name(model, character_lookup, max_length)
```

## Model 2 - Use whether it's a cat or a dog in the name generation

This is the same model as 1, except now we want to have the names depend
on if it's a cat or a dog. The training data is now previous characters in the
name AND the species of animal. This requires a more complex network.


Generate the X data now requires keeping track of the species.

```{r model_2_generate_data}
subsequence_data_v2_raw <-
  pet_data %>%
  filter(species %in% c("cat","dog")) %>%
  mutate(accumulated_name =
           name %>%
           str_c("+") %>%
           str_split("") %>%
           map( ~ purrr::accumulate(.x,c))
         ) %>%
  select(species, accumulated_name) %>%
  unnest(accumulated_name)

subsequence_data_v2 <- subsequence_data_v2_raw$accumulated_name
species_data_v2 <- (subsequence_data_v2_raw$species == "cat")*1L
```

There are now two (!) X matrices, one for previous name and one for species.

```{r model_2_generate_data_continued}
text_matrix_v2 <-
  subsequence_data_v2 %>%
  map(~ character_lookup$character_id[match(.x,character_lookup$character)]) %>% 
  pad_sequences(maxlen = max_length+1) %>% 
  to_categorical(num_classes = num_characters) 

x_name_v2 <- text_matrix[,1:max_length,]
x_species_v2 <- species_data_v2
y_name_v2 <- text_matrix[,max_length+1,]
```

The network is now more complex with two inputs and a concatenate.

```{r model_2_define_model}

# we'll fill this in

```

The name generation function needs to be adjusted to account for the species too, but that's
left as an exercise for the reader!

### Model 3 - predict if the name is for a cat or a dog

Instead of trying to predict the next character, let's take a name and predict
if it's a cat or dog's name. The things that need to change are:

1. The y matrix is now the species
2. The names should be padded on the right not the left

```{r model_3_generate_data}

x_max_length <- 20

prediction_data <-
  pet_data %>% 
  filter(species %in% c("cat","dog")) %>%
  select(name, species)

pred_text_matrix <-
  prediction_data %>%
  pull(name) %>%
  str_split("") %>%
  map(~ character_lookup$character_id[match(.x,character_lookup$character)]) %>% 
  pad_sequences(maxlen = x_max_length, padding = "post", truncating = "post") %>% 
  to_categorical(num_classes = num_characters) 

pred_class <-
  as.array((prediction_data$species == "cat")*1L) # needs to be a number!

```

The activation now has to change to sigmoid because we are trying to get a 0-1 probability

```{r model_3_define_model}

# we'll fill this in

```

Model fitting is the same as the other models.

```{r model_3_fit_model}
fit_results <- pred_model %>% 
  fit(
    pred_text_matrix, 
    pred_class,
    batch_size = 64,
    epochs = 16,
    validation_split = 0.05
)
```

We can then run the model on some names to see what happens.

```{r model_3_prediction}
probability_cat <- function(example_name){
  formatted_example_name <- example_name %>%
    str_split("") %>%
    map(~character_lookup$character_id[match(.x,character_lookup$character)]) %>%
    pad_sequences(maxlen = x_max_length) %>%
    to_categorical(num_classes = num_characters)

  # we must account for the fact that it returns a matrix not a single number
  predict(pred_model, formatted_example_name)[1,1]
}

many_names <- pet_data %>% 
  sample_n(100) %>% 
  select(name, species) %>% 
  mutate(probability_cat = map_dbl(name, probability_cat))
```


## Next steps

* Predict text at the word level instead of the character level: use an encoding layer
* Use a model that was already trained and adjust it to your data: transfer learning
* Train faster with a GPU instead of a CPU: That's easy to do with Saturn Cloud, see the talk Intro to Neural Networks in R: https://saturncloud.io/events/
* Learn more by reading Deep Learning with R: https://www.manning.com/books/deep-learning-with-r