---
title: Futures in R
description: An overview of the R future package
output: html_notebook
---

The [future](https://cran.r-project.org/web/packages/future/index.html) package is a simple and uniform method for asynchonous computing. It provides a way of submitting functions that don't block the current R session. It is cross-platform and works on a variety of backends.

A future is an object that represents a promise to return the value of a function when it is computed. The future itself executes almost instantly, but the function continues to process in the background on a separate R process. This delegation of execution allows the main R process to move on to other computations while the background process continues to execute. 

A future consists of three parts:

1. The execution environment
2. The expression to the executed
3. The status of the execution

## The Execution Environment

To start using futures, first specify the execution environment. The execution environment determines where to execute the calculation defined by the future. It is defined using the `plan` function. Consider this example:

```{r plan example, eval = FALSE}
plan(multisession)
```

> **Note**: The result of a future does not depend on what `plan` you choose; a future behaves consistently across plans. 

There are several built-in plans for local and distributed computing:

* **sequential**: A non-parallel, blocking execution environment -- This is the standard `plan`. Think of this as the default single-threaded R process. You won't see any performance changes by using futures in this environment, but this `plan` can be useful for debugging and testing.
* **multisession**: A parallel, non-blocking execution environment -- This `plan` will allow futures to utilize additional background R sessions to perform computations, which is useful on local machines where you want to utilize more than one core.
* **multisession**: A parallel, non-blocking execution environment -- Instead of running on background R sessions, like multisession, this `plan` runs on forked R processes. This approach has some speed advantages, particularly a reduction in memory overhead, but it is not stable on Windows or RStudio. 
* **cluster**: A parallel, non-blocking execution environment -- It allows futures evaluations to run on external R sessions. It can be used on both local and remote machines. 

In addition to these built-in plans, futures can also be used with [batchtools](https://github.com/mllg/batchtools), [callr](https://cran.r-project.org/web/packages/future.callr/index.html), or custom backends.

## The Expression to Be Executed
Futures can be created either implicitly or explicity. Both styles work to the same end but with different syntax: implict futures are more similar to regular R code, whereas explicit futures can be clearer and easier to read. 

To define an implicit future, simply replace `<-` with `%<-%`, like in the example below. The value of an implicitly defined future is returned automatically.

```{r implict future example, eval = FALSE}
v %<-% {
    expression
}
```


To define an explicit future, wrap it in the `future` function. For explict futures, you need to request the value manually. Consider the example below:

```{r explicit future example, eval = FALSE}
f <- future({
    expression
})
v <- value(f)
```

We will use explict futures for the remainder of this document.

## The Status of the Execution

The state of a future can be either resolved or unresolved -- either finished with its calculation or still running. Calling `value()` will automatically block the main R process until the future is resolved and then return the value. If you want to check whether a future is resolved without blocking, you can use the `resolved` function, shown below:

```{r resolved example, eval = FALSE}
f <- future({
    expression
}) # create the future

r <- resolved(f) # check if it is done

v <- value(f) # wait and get the result
```

## Let's Look at Some Examples

Let's start by defining some simple functions:
```{r imports and functions}
library(future)

increment <- function(x) {
    Sys.sleep(runif(1))
    return(x + 1)
}

double <- function(x) {
    Sys.sleep(runif(1))
    return(2 * x)
}

add <- function(x, y) {
    Sys.sleep(runif(1))
    return(x + y)
}
```


Let's try out the sequential environment. We define the seed parameter in the future function due to the use of random numbers in the functions:

```{r sequential plan}
plan(sequential)

future <- future(increment(1), seed = NULL)

resolved(future)
print(value(future))
```

As you can see, `resolved()` returned TRUE, showing that the future was evaluated completely before the R process moved on to the `resolved` function. This is exactly how we would expect the function to work without the futures package.

If we want to have non-blocking code, we need to use one of the other environments. Here we use `multisession`:

```{r mulitisession plan}
plan(multisession)

future <- future(increment(1), seed = TRUE)

resolved(future)
print(value(future))
```

This time, the `resolved` function returned FALSE. This shows that, instead of waiting until the future completed, the R process continued on before finally waiting for the function to complete at the `value` function. 

Note that, although futures can be nested, they can't be chained. Before using a future as an input to another future, you need to evaluate and get its values, as shown below:

```{r future from futures explicit}
x <- future(increment(1), seed = TRUE)
y <- future(double(2), seed = TRUE)

# evaluate and get the values of the futures
x <- value(x)
y <- value(y)

# pass them to the next future
z <- future(add(x, y), seed = TRUE)

resolved(z)
print(value(z))
```

Before ending your session, revert the plan to sequential computations:
```{r sequential plan reset}
plan(sequential)
```

## Conclusion
The future package is an excellent tool for parrallelizing code in R. It allows for the re-use of code on a variety of backends, including on clusters. It also serves as the backend for other parallel packages like [furrr](https://github.com/DavisVaughan/furrr) or [targets](https://docs.ropensci.org/targets/).

If you want to learn more about the future package, be sure to read the [vingnettes](https://cran.r-project.org/web/packages/future/index.html) published for the package.
